Precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-17 manuscript.tex
3.1

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Precisely timing motifs in computational neuroscience}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In the following section, we will review theoretical models which directly tackle the problem of using precise spiking motifs. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
A natural candidate to use these precise temporal patterns in the brain is to use Spiking Neural Networks (SNNs)~\citep{maass_networks_1997}. The approach which is at present most prominent in the SNNs community is to use existing algorithms from machine learning and to adapt them to the specificity of spiking architectures. One such example is to adapt the successes of deep learning algorithms and to transfer the back-propagation algorithm to SNNs, for instance with a surrogate gradient and try to ``cross-compile'' classical Neural Network to a spiking architecture. This approach is quite successful, and SNNs approach in some case the performance of Deep Learning algorithms, for instance on the N-MNIST dataset for categorizing digits in a stream of events. However, most biological neural systems use spikes and are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption. There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs.  To go beyond the state-of-the-art, we will review here on one core computation of a spiking neuron, that is, is its ability to switch from the classical integrator mode (summing analog currents on its synapses) to a synchrony detector where it emits a spike whenever presynaptic spiking inputs are synchronized. To overcome the diversity of input presynaptic patterns, we will explore different existing architectures to learn to detect stable ``polychronous`` events, that is, volleys of spikes which are stable up to certain synaptic delays. These models will be compared in light of neuroscientific and computational perspectives. We review theoretical and computational foundations of PG detection in models. In particular, our objective is to describe methods which fully exploit the capacity of spiking neurons to detect synchronous patterns.

Finally we will discuss future avenues for effective PG detection and learning in event streams.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Izhikevitch's polychronization model}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

First, we will review models for the detection of polychronous groups:

Our goal in this review is to bring an interdisciplinary perspective on the computational advantage of time series representations for the brain and for information processing machines. In particular, we will focus on the hypothesis that there exists at the scale of an assembly of neurons a representation based on a set of motifs defined by different relative spike times. %Here, we will review current literature on the detection of such motifs in generic raster plots. It is work in progress, where anybody interested can \emph{openly} join.
%We will first review some biological and theoretical evidence in neural information processing in the particular case of vision. 
%The next section will focus on precise spike time motifs. 
We will then present some models for the detection of such motifs in arbitrary raster plots, biological or artificial (notably from event-based cameras). In particular, we will discuss models which exploit the variety of synaptic delays on the dendritic tree. Then, we will try to outline some possible strategies for learning these patterns and finally discuss possible perspectives.

@@

A popular model for the detection of latency patterns is the tempotron~\citep{gutig_tempotron_2006}. This model is in particular reviewed in~\citep{gutig_spike_2014}. The Tempotron is a supervised synaptic learning algorithm which classifies a distractor from a target motif. This aims at extending the perceptron which does not incorporate a spike timing framework. The tempotron learning rule is derived by an optimization process and takes the form of a STDP rule. It is general consensus that spike timing (STDP) plays a crucial role in the development of synaptic efficacy for many different kinds of neurons~\citep{caporale_spike_2008}. The limits of this model is that it's output is only binary and that it's storage capacity are limited.

@@

% polychronization
%  blog post by Paxon Frady \url{https://epaxon.blogspot.com/2012/07/izhikevich-2006-polychronization.html} Basic idea: different temporal activation of neurons can activate different populations based on conduction delays. This allows for a much higher dimensional representation space, and a lot more memory capacity. Firing is not synchronous, but it is time-locked = polychrony. STDP automatically associates the neurons in the network into polychronous groups. There can be far more polychronous groups than neurons. He lets the network just settle for 24 hours - and these groups emerge spontaneously from the random connectivity initial conditions. The dendritic tree can be used to select the polychronous inputs more precisely. Can act as a full feed-forward neural network and make non-linear classifications. Two dendritic trees on pyramidal cells - one for the feed-forward input, and one for feed-back/recurrent. These can interact such that the feed-back connections are predicting the feed-forward inputs.

Considering the number of spikes required to exceed a voltage threshold, asynchronous signals are less efficient than synchronous signals. However, taking axonal propagation times into account, synchronous signals may not coincide significantly at the post-synaptic neuron contrary to asynchronous or polychronous signals. The term polychronous was first introduced in 2006, by E. Izhichevitch in~\citep{izhikevich_polychronization_2006}. He defines this term after pointing out a particular organization of the neurons of his spiking artificial neural network. The network is characterized by a timing-dependant learning rule for weights (STDP) and by fixed conduction delays between neurons. Due to the interplay between the delays and STDP, the spiking neurons spontaneously self-organize into groups and generate patterns of stereotypical polychronous activity, i.e.~exhibit reproducible time-locked but not synchronous firing patterns. The neurons composing a group discharge at different times, but due to delays, the spikes reach the postsynaptic neuron at the same time. This synchrony leads to the summation of the excitatory post-synaptic potentials evoked by each spike and thus to the crossing of the voltage threshold and to the discharge of a spike. According to the STDP rule, the neurons involved in this activity will see their weight of synaptic connection increase and thus, constitute a polychronous group. Interestingly, thanks to the fact that a neuron can be involved in different polychronous groups, the number of coexisting polychronous groups far exceeds the number of neurons in the network, resulting in an unprecedented memory capacity of the system.

This paper proposes a supervised delay learning algorithm for spiking neurons with temporal encoding, in which both the weight and delay of a synaptic connection can be adjusted to enhance the learning performance~\citep{wang}.

Thus, the learning of delays allowing this polychronous group organization may be useful to detect temporal sequences of interest.

Wang et al. (2013) presented a hardware implementation of polychronous networks in which propagation delays are learned in a supervised manner, based on the expected firing time of the post-synaptic neuron.

Reproducing Polychronization: A Guide to Maximizing the Reproducibility of Spiking Network Models \url{https://www.frontiersin.org/articles/10.3389/fninf.2018.00046/full}     comes with python code

    The recent ``multi-neuronal spike sequence detector'' (MNSD) architecture integrates the weight- and delay-adjustment methods by combining plasticity with the neurocomputational feature spike latency : \url{https://pubmed.ncbi.nlm.nih.gov/33679293/}
    an extensive (graph-centric) review on \href{https://arxiv.org/abs/2109.07618}{Synchronization in time-varying networks}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Precisely timing motifs in computational neuroscience}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In the following section, we will review theoretical models which directly tackle the problem of using precise spiking motifs. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
A natural candidate to use these precise temporal patterns in the brain is to use Spiking Neural Networks (SNNs)~\citep{maass_networks_1997}. The approach which is at present most prominent in the SNNs community is to use existing algorithms from machine learning and to adapt them to the specificity of spiking architectures. One such example is to adapt the successes of deep learning algorithms and to transfer the back-propagation algorithm to SNNs, for instance with a surrogate gradient and try to ``cross-compile'' classical Neural Network to a spiking architecture. This approach is quite successful, and SNNs approach in some case the performance of Deep Learning algorithms, for instance on the N-MNIST dataset for categorizing digits in a stream of events. However, most biological neural systems use spikes and are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption. There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs.  To go beyond the state-of-the-art, we will review here on one core computation of a spiking neuron, that is, is its ability to switch from the classical integrator mode (summing analog currents on its synapses) to a synchrony detector where it emits a spike whenever presynaptic spiking inputs are synchronized. To overcome the diversity of input presynaptic patterns, we will explore different existing architectures to learn to detect stable ``polychronous`` events, that is, volleys of spikes which are stable up to certain synaptic delays. These models will be compared in light of neuroscientific and computational perspectives. We review in this section theoretical and computational foundations of PG detection in models. In particular, our objective is to describe methods which fully exploit the capacity of spiking neurons to detect synchronous patterns.
%Finally we will discuss future avenues for effective PG detection and learning in event streams.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Izhikevitch's polychronization model}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
First, we will review models for the detection of polychronous groups.
% Our goal in this review is to bring an interdisciplinary perspective on the computational advantage of time series representations for the brain and for information processing machines. 
In particular, we will focus on the hypothesis that there exists at the scale of an assembly of neurons a representation based on a set of motifs defined by different relative spike times. %Here, we will review current literature on the detection of such motifs in generic raster plots. It is work in progress, where anybody interested can \emph{openly} join.
%We will first review some biological and theoretical evidence in neural information processing in the particular case of vision. 
%The next section will focus on precise spike time motifs. 
We will then present some models for the detection of such motifs in arbitrary raster plots, biological or artificial (notably from event-based cameras). In particular, we will discuss models which exploit the variety of synaptic delays on the dendritic tree. Then, we will try to outline some possible strategies for learning these patterns and finally discuss possible perspectives.
%  blog post by Paxon Frady \url{https://epaxon.blogspot.com/2012/07/izhikevich-2006-polychronization.html} Basic idea: different temporal activation of neurons can activate different populations based on conduction delays. This allows for a much higher dimensional representation space, and a lot more memory capacity. Firing is not synchronous, but it is time-locked = polychrony. STDP automatically associates the neurons in the network into polychronous groups. There can be far more polychronous groups than neurons. He lets the network just settle for 24 hours - and these groups emerge spontaneously from the random connectivity initial conditions. The dendritic tree can be used to select the polychronous inputs more precisely. Can act as a full feed-forward neural network and make non-linear classifications. Two dendritic trees on pyramidal cells - one for the feed-forward input, and one for feed-back/recurrent. These can interact such that the feed-back connections are predicting the feed-forward inputs. 
Considering the number of spikes required to exceed a voltage threshold, asynchronous signals are less efficient than synchronous signals. However, taking axonal propagation times into account, synchronous signals may not coincide significantly at the post-synaptic neuron contrary to asynchronous or polychronous signals. The term polychronous was first introduced by~\citet{izhikevich_polychronization_2006}. He defines this term after pointing out a particular organization of the neurons of his spiking artificial neural network. The network is characterized by a timing-dependant learning rule for weights (STDP) and by fixed conduction delays between neurons. Due to the interplay between the delays and STDP, the spiking neurons spontaneously self-organize into groups and generate patterns of stereotypical polychronous activity, i.e.~exhibit reproducible time-locked but not synchronous firing patterns. The neurons composing a group discharge at different times, but due to delays, the spikes reach the postsynaptic neuron at the same time. This synchrony leads to the summation of the excitatory post-synaptic potentials evoked by each spike and thus to the crossing of the voltage threshold and to the discharge of a spike. According to the STDP rule, the neurons involved in this activity will see their weight of synaptic connection increase and thus, constitute a polychronous group. Interestingly, thanks to the fact that a neuron can be involved in different polychronous groups, the number of coexisting polychronous groups far exceeds the number of neurons in the network, resulting in an unprecedented memory capacity of the system.

% polychronization
Interestingly the paper by~\citet{izhikevich_polychronization_2006} stirred a debate in the field of computational neuroscience, with a general postive acceptance, but quite few works extended this seminal paper.

@@

A popular model for the detection of latency patterns is the tempotron~\citep{gutig_tempotron_2006}. This model is in particular reviewed in~\citep{gutig_spike_2014}. The Tempotron is a supervised synaptic learning algorithm which classifies a distractor from a target motif. This aims at extending the perceptron which does not incorporate a spike timing framework. The tempotron learning rule is derived by an optimization process and takes the form of a STDP rule. It is general consensus that spike timing (STDP) plays a crucial role in the development of synaptic efficacy for many different kinds of neurons~\citep{caporale_spike_2008}. The limits of this model is that it's output is only binary and that it's storage capacity are limited.
The recent ``multi-neuronal spike sequence detector'' (MNSD) architecture integrates the weight- and delay-adjustment methods by combining plasticity with the neurocomputational feature spike latency : \url{https://pubmed.ncbi.nlm.nih.gov/33679293/}
an extensive (graph-centric) review on \href{https://arxiv.org/abs/2109.07618}{Synchronization in time-varying networks}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Extensions of Izhikevitch's polychronization model}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A notable exception is an extensive work aiming at reproducing the polychronization model~\citep{pauli_reproducing_2018}. Indeed, the original paper contained material within the text to reproduce the whole model. This more recent work details how this code was slighlty corrected. It comes with python code and a version control system detailing the whole process used to give provenance to the different steps in this scientific process.

@@

@@

 ~\citet{wang_neuromorphic_2015} presented a hardware implementation of polychronous networks in which propagation delays are learned in a supervised manner, based on the expected firing time of the post-synaptic neuron. This paper proposes a supervised delay learning algorithm for spiking neurons with temporal encoding, in which both the weight and delay of a synaptic connection can be adjusted to enhance the learning performance.
